Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2022-12-28 manuscript.tex

% define document type (i.e., template. Here: A4 APA manuscript with 12pt font)
\documentclass[man, 12pt, a4paper, mask]{apa7}

@@

Recent years have seen a striking increase in the number and variety of research studies that follow participants' everyday experiences and collect real-world psychological time series \citep[e.g., experience sampling method (ESM),][; also see \fgrref{fig:ScopusEsm}]{hamaker2017}. These extensive longitudinal data sets come with new forms of heterogeneity, where researchers have to consider differences across large numbers of participants, time points, and variables \citep[e.g.,][]{cattell1966, Monden2015, wardenaar2013}. Oftentimes, however, researchers are interested in precisely this complexity and wish to understand how people differ in their temporal developments across a number of distinct variables \citep[i.e., understand the structure of between-subject differences of within-participant dynamics; e.g.,][]{ernst2021}. One relatively recent addition to the toolbox of ESM researchers has been time series clustering --- the idea of grouping participants based on the characteristics of their time series dynamics \citep{ariens2020}. In practice, this means that participants are inductively grouped in such a way that the dynamics of some variable(s) have maximum similarity for individuals in the same group, and minimum similarity for individuals in different groups \citep[e.g.,][]{liao2005}.

Time series clustering has a number of conceptual use cases. Specifically, time series clustering aids in reducing the time, variable, and person complexity by extracting structure between participants. These reduction and structuring qualities can be essential in detecting phenomena and extracting more abstract functional principles \citep[][]{eronen2021a}. Such meaningful psychological principles and phenomena have often been recognized as being imperative for building robust theories and real-world interventions \citep[e.g.,][]{borsboom2021, eronen2020}.

@@

\subsection{Features of Psychological Time Series}

\paragraph{Inertia and Periodicity}

\paragraph{Analysis Steps}

sort of feature-based clustering in psych \citep{heylen2016}

@@

@@

\begin{figure}[hbtp]
  \caption{Scopus ESM Development}
  \label{fig:ScopusEsm}
  \centering\includegraphics[width=\textwidth]{figures/Scopus-ESM-Development.png}
  \caption*{Note: \\
  Search Terms: `( "ecological momentary assessment"  OR  "experience sampling"  OR  "intensive longitudinal")' in Title, Abstract, or Keyword.}
\end{figure}

\begin{figure*}[hbtp]
  \caption{Flowchart Feature-Based Time Series Clustering in Psychology}
  \label{fig:TSCFlow}
  \centering\includegraphics[width=\textwidth]{figures/TS Cluster Flow/TimeSeriesClusterFlowSelection.pdf}
  \caption*{Note: \\
  Choices selected for illustration in this manuscript are marked in bold.}
\end{figure*}

\begin{figure*}[hbtp]
  \caption{Exemplary Flowchart of Data Points in Feature-Based Time Series Clustering}
  \label{fig:TSCFlowN}
  \centering\includegraphics[width=\textwidth]{figures/TS Cluster Flow/tsClustFlowN.pdf}
  \caption*{Note: \\
  The presented number of participants, variables, and measurement occasions are somewhat arbitrary but generally represent common sample sizes found within the literature. Also the number of extracted clusters is presented for illustrative purposes only.}
\end{figure*}

\begin{figure*}[hbtp]
  \caption{Cluster Group Comparisons based on Features and Variables}
  \label{fig:clusterFeatVar}
  \centering\includegraphics[width=\textwidth]{figures/clusterFeatVarComb.pdf}
  \caption*{Note: \\
  Within the ``(B) Feature Focus'' subplot, the `n' and `Discrimination' comparison variables were not part of the original time series clustering.}
\end{figure*}

\begin{figure*}[hbtp]
  \caption{Cluster Group Comparisons over time}
  \label{fig:clusterTs}
  \centering\includegraphics[width=\textwidth]{figures/clusterTsComb.pdf}
  \caption*{Note: \\
  Subplot (A) displays the variable cluster means at every measurement occasion. Subplot (B) shows the GAM spline for each cluster across the measurement occasions.}
\end{figure*}

>>>>

% define document type (i.e., template. Here: A4 APA manuscript with 12pt font)
\documentclass[man, 12pt, a4paper, mask, floatsintext]{apa7}

@@

Recent years have seen a striking increase in the number and variety of research studies that follow participants' everyday experiences and collect real-world psychological time series \citep[e.g., experience sampling method (ESM),][; also see \fgrref{fig:ScopusEsm}]{hamaker2017}. These extensive longitudinal data sets come with new forms of heterogeneity, where researchers have to consider differences across large numbers of participants, time points, and variables \citep[e.g.,][]{cattell1966, Monden2015, wardenaar2013}. Oftentimes, however, researchers are interested in precisely this complexity and wish to understand how people differ in their temporal developments across several variables \citep[i.e., understand the structure of between-subject differences of within-participant dynamics; e.g.,][]{ernst2021}. One relatively recent addition to the toolbox of ESM researchers has been time series clustering --- the idea of grouping participants based on the similarity of their time series \citep{ariens2020}.

\begin{figure}[!hbtp] %hbtp
  \caption{Scopus ESM Development}
  \label{fig:ScopusEsm}
  \centering\includegraphics[width=\textwidth]{figures/Scopus-ESM-Development.png}
  \caption*{Note: \\
  Search Terms: `( "ecological momentary assessment"  OR  "experience sampling"  OR  "intensive longitudinal")' in Title, Abstract, or Keyword.}
\end{figure}

The main aim of time series clustering is to inductively group participants in such a way that the dynamics of some variable(s) have maximum similarity for individuals in the same group, and minimum similarity for individuals in different groups \citep[e.g.,][]{liao2005}. In practice, this means time series clustering critically depends on identifying similarities or dissimilarities in time series and then accurately distinguishing different groups (REF).

1. phenomena and principle detection
    a. differences between clusters
    b. patterns within clusters (\~ early warning signals?)
2.

Time series clustering has a number of conceptual use cases with psychological data. Prime among them is the ability to reduce the time, variable, and person complexity by extracting structure between participants. These reduction and structuring qualities can be essential in detecting phenomena and extracting more abstract functional principles \citep[][]{eronen2021a}. 
These phenomena and principles can be (1) meaningful differences between clusters, as well (2) important patterns, trends, and relationships within clusters \citep[e.g.,][]{schrodt2000}.

Once distinct groups and patterns have been identified, researchers can examine the extent to which these within-group and between-group structures are associated with other variables of interest, such as personality traits, demographic characteristics, or other psychological constructs (REFERENCE).

Such meaningful psychological principles and phenomena have often been recognized as being imperative for building robust theories and real-world interventions \citep[e.g.,][]{borsboom2021, eronen2020}. 
Additionally, time series clustering has been used to inform the development of targeted interventions and therapies for individuals, for example, with mood disorders and other psychological conditions (REFERENCE).

@@

\subsection{Features of Psychological Time Series}
There are hundreds of possible ts features

\citet{kuppens2017} originally proposed (1) variability, (2) co-variance or ICC, (3) inertia, and (4) cross-lagged correlations. These features were then extended and formalized by \citet{krone2018} into (1) within-person variance, (2) innovation, (3) inertia or autocorrelation, (4) cross-lagged regression parameters, (5) covariances as granularity, and (6) mean intensity. \citet{krone2018} were even able to use these features to cluster study participants.

\citet{dejonckheere2019} discuss (1) mean, (2) variance, (3) relative variance, (4) instability, (5) inertia / autoregression, (6) interdependence / network density, (7) granularity or differentiation / ICC, (8) correlation, and (9) diversity / Gini coefficient.

Outside the psychological literature, time series features have also included more summaries in terms of measurement distribution, categorical states, and, importantly, nonlinear developments.

\citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data.

\citet{wang2006} propose a set of nine statistical features for describing a trajectory: (1) trend, (2) seasonality, (3) periodicity, (4) skewness, (5) kurtosis, (6) serial correlation, (7) non-linearity, (8) self-similarity, and (9) chaos. This selection was also used by \citep{fulcher2013}. \citet{adya2001} proposed a total of 28 features relevant for forecasting, which largely fit within the selection proposed by \citet{wang2006}.

@@

\paragraph{Inertia and Periodicity} self-similarity, seasonality, periodicity, inertia

\subsection{Analysis Steps}

parameter-based clustering: VAR \citep{ernst2021} splines \citep{axen2011}

sort of feature-based clustering in psych \citep{heylen2016} as well as \citep{krone2018}

\begin{figure}[!ht] % \begin{figure*}[hbtp] <-- on its own page
  \caption{Flowchart Feature-Based Time Series Clustering in Psychology}
  \label{fig:TSCFlow}
  \centering\includegraphics[width=\textwidth]{figures/TS Cluster Flow/TimeSeriesClusterFlowSelection.pdf}
  \caption*{Note: \\
  Choices selected for illustration in this manuscript are marked in bold.}
\end{figure}

\begin{figure}[!ht] %hbtp
  \caption{Exemplary Flowchart of Data Points in Feature-Based Time Series Clustering}
  \label{fig:TSCFlowN}
  \centering\includegraphics[width=\textwidth]{figures/TS Cluster Flow/tsClustFlowN.pdf}
  \caption*{Note: \\
  The presented number of participants, variables, and measurement occasions are somewhat arbitrary but generally represent common sample sizes found within the literature. Also the number of extracted clusters is presented for illustrative purposes only.}
\end{figure}

@@

\begin{figure*}[!ht] %hbtp
  \caption{Cluster Group Comparisons based on Features and Variables}
  \label{fig:clusterFeatVar}
  \centering\includegraphics[width=\textwidth]{figures/clusterFeatVarComb.pdf}
  \caption*{Note: \\
  Within the ``(B) Feature Focus'' subplot, the `n' and `Discrimination' comparison variables were not part of the original time series clustering.}
\end{figure*}

\begin{figure*}[!ht] %hbtp
  \caption{Cluster Group Comparisons over time}
  \label{fig:clusterTs}
  \centering\includegraphics[width=\textwidth]{figures/clusterTsComb.pdf}
  \caption*{Note: \\
  Subplot (A) displays the variable cluster means at every measurement occasion. Subplot (B) shows the GAM spline for each cluster across the measurement occasions.}
\end{figure*}
